---
title: "GSE193979_2 TB Diagnosis, 2 Month, and 6 Month Control Data"
output: html_document 
date: "2025-02-22"
---

Load data
```{r}
library(data.table)
tpm_data <- read.delim("GSE193979_norm_counts_TPM_GRCh38.p13_NCBI.tsv", row.names = 1, check.names = FALSE)
```

Select the columns that you want to keep - diagnosis sample, 2 month samples, 6 month samples for people who are TB Only 
```{r}
selected_ids <- c("GSM5824894", "GSM5824838", "GSM5824984", "GSM5824932", "GSM5824970", "GSM5824979", "GSM5824847", "GSM5824855", "GSM5824943", "GSM5824904", "GSM5824967", "GSM5824885", "GSM5824852", "GSM5824839", "GSM5824858", "GSM5824878", "GSM5824868", "GSM5824955")
```

Filter the data to keep only the selected samples defined above: 
```{r}
filtered_tpm <- tpm_data[, colnames(tpm_data) %in% selected_ids]
```

Save the filtered data as csv:
```{r}
write.csv(filtered_tpm, "filtered_tpm.csv", quote = FALSE)
```

Load/install packages to convert Entrez IDs to HGNC IDs:
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")}

# Force reinstall both packages
BiocManager::install(c("AnnotationDbi", "org.Hs.eg.db"), force = TRUE)

# Reload the libraries
library(AnnotationDbi)
library(org.Hs.eg.db)
```
Test select() function on a test set:
```{r}
test_ids <- c("100287102", "653635", "102466751")  # A few example Entrez IDs

# Ensure org.Hs.eg.db is correctly loaded
class(org.Hs.eg.db)  # Should return "OrgDb"

# Run select() on a small test set
test_mapping <- select(org.Hs.eg.db, keys = test_ids, keytype = "ENTREZID", columns = c("SYMBOL"))
print(test_mapping)
```

Map Entrez IDs to HGNC Symbols to get 1:1 mapping:
```{r}
# Extract Entrez IDs from filtered_tpm
entrez_keys <- rownames(filtered_tpm)

# Convert Entrez IDs to HGNC Symbols
entrez_to_hgnc <- select(org.Hs.eg.db, keys = entrez_keys, keytype = "ENTREZID", columns = c("SYMBOL"))

# Remove NAs and duplicates
entrez_to_hgnc <- na.omit(entrez_to_hgnc)
entrez_to_hgnc <- entrez_to_hgnc[!duplicated(entrez_to_hgnc$ENTREZID), ]

# Create a named vector for mapping
entrez_to_hgnc_vector <- setNames(entrez_to_hgnc$SYMBOL, entrez_to_hgnc$ENTREZID)

# Apply mapping (replace Entrez IDs with HGNC symbols)
filtered_tpm$HGNC <- entrez_to_hgnc_vector[rownames(filtered_tpm)]
```

Fix duplicate HGNC ID Symbols (aggregate by mean):
```{r}
library(dplyr)

# Convert rownames to characters (ensure no format issues)
rownames(filtered_tpm) <- as.character(rownames(filtered_tpm))

# Create a named vector for mapping Entrez to HGNC
entrez_to_hgnc_vector <- setNames(entrez_to_hgnc$SYMBOL, entrez_to_hgnc$ENTREZID)

# Apply mapping (replace Entrez IDs with HGNC symbols)
filtered_tpm$HGNC <- entrez_to_hgnc_vector[rownames(filtered_tpm)]

# Remove genes that didn't map
filtered_tpm <- filtered_tpm[!is.na(filtered_tpm$HGNC) & filtered_tpm$HGNC != "", ]

# Aggregate duplicate HGNC symbols by averaging TPM values
filtered_tpm <- filtered_tpm %>%
  group_by(HGNC) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  ungroup()

# Convert back to matrix
filtered_tpm <- as.data.frame(filtered_tpm)
rownames(filtered_tpm) <- filtered_tpm$HGNC
filtered_tpm$HGNC <- NULL  # Remove redundant column

# Final check: Ensure no duplicates
sum(duplicated(rownames(filtered_tpm)))  # Should return 0
```

Check to make sure everything looks okay:
```{r}
# Ensure no duplicates exist
sum(duplicated(rownames(filtered_tpm)))  # Should return 0

# Verify that row names are now HGNC symbols
head(rownames(filtered_tpm))  # Should return HGNC gene names and not Entrez IDs

# Save the final dataset as csv for future use
write.csv(filtered_tpm, "filtered_tpm_with_HGNC.csv", quote = FALSE, row.names = TRUE)
```

Define Sample Metadata to convert the data into a SummarizedExperiment with the three time points: 
```{r}
library(SummarizedExperiment)

metadata <- data.frame(
  Sampler = rep(c("A", "B", "C", "D", "E", "F"), each = 3),  # Sampler label
  SampleID = c("GSM5824894", "GSM5824847", "GSM5824858",  # A
               "GSM5824838", "GSM5824855", "GSM5824839",  # B
               "GSM5824984", "GSM5824943", "GSM5824868",  # C
               "GSM5824932", "GSM5824904", "GSM5824955",  # D
               "GSM5824970", "GSM5824967", "GSM5824852",  # E
               "GSM5824979", "GSM5824885", "GSM5824878"), # F
  Timepoint = rep(c("Diagnosis", "Month2", "Month6"), times = 6))

# Set row names as Sample IDs
rownames(metadata) <- metadata$SampleID
metadata$SampleID <- NULL
```

Ensure filtered_tpm mataches the metadata dataframe:
```{r}
# Filter TPM data to include only the selected samples
filtered_tpm <- filtered_tpm[, colnames(filtered_tpm) %in% rownames(metadata)]

# Ensure TPM data columns match the metadata order
filtered_tpm <- filtered_tpm[, rownames(metadata)]
```

Create and verify the SummarizedExperiment (se)
```{r}
se <- SummarizedExperiment(
  assays = list(TPM = as.matrix(filtered_tpm)),  # TPM expression matrix
  colData = metadata  # Sample metadata with Sampler and Timepoint
)

dim(se)  # Should return 37691 genes x 18 samples (6 samplers x 3 timepoints)
colData(se)  # Should display Sampler IDs and Timepoints correctly
head(assay(se))  # Should show TPM values for genes
```
Check the structure of the loaded data
```{r}
str(se)
```

Load PI3K, TGF Beta, P53, Notch, and WNT-Beta Catenin gene sets:
```{r}
Pi3k_akt_gmt_file <- ("~/TBLung/TB_6Month/HALLMARK_PI3K_AKT_MTOR_SIGNALING.v2024.1.Hs.gmt")
TGF_BETA_gmt_file <- ("~/TBLung/TB_6Month/HALLMARK_TGF_BETA_SIGNALING.v2024.1.Hs.gmt")
P53_gmt_file <- ("~/TBLung/TB_6Month/HALLMARK_P53_PATHWAY.v2024.1.Hs.gmt") 
Notch_gmt_file <- ("~/TBLung/TB_6Month/HALLMARK_NOTCH_SIGNALING.v2024.1.Hs.gmt")
WNT_BETA_CATENIN_gmt_file <- ("~/TBLung/TB_6Month/HALLMARK_WNT_BETA_CATENIN_SIGNALING.v2024.1.Hs.gmt")
```

Load the gene sets using getGmt()
```{r}
library(readr)
library(GSEABase)

pi3k_akt_genesets <- getGmt(Pi3k_akt_gmt_file)
TGF_BETA_genesets <- getGmt(TGF_BETA_gmt_file )
p53_genesets <- getGmt(P53_gmt_file)
Notch_gmt_genesets <- getGmt(Notch_gmt_file)
WNT_BETA_CATENIN_genesets <- getGmt(WNT_BETA_CATENIN_gmt_file )

# View the first few pathways and genes
head(pi3k_akt_genesets)
head(TGF_BETA_genesets)
```

Load TBSignatureProfiler
```{r}
library(TBSignatureProfiler)
```

Add gene sets to TBsignatures:
```{r}
TBsignatures$'HALLMARK_PI3k_akt' <-pi3k_akt_genesets
TBsignatures$'HALLMARK_TGF_BETA' <- TGF_BETA_genesets
TBsignatures$'HALLMARK_P53' <- p53_genesets
TBsignatures$'HALLMARK_NOTCH' <- Notch_gmt_genesets
TBsignatures$'HALLMARK_WNT_BETA_CATENIN' <- WNT_BETA_CATENIN_genesets
```

Select the signagtures you want to view:
```{r}
sample_tbsignatures <- TBsignatures
names(sample_tbsignatures)

#selecting the "Sweeny_OD_3", "Zak_RISK_16", "HALLMARK_P53", "HALLMARK_PI3k_akt", "HALLMARK_TGF_BETA", "HALLMARK_WNT_BETA_CATENIN"
sample_tbsignatures3 <- sample_tbsignatures[c(64,79,81,82,83,85)]
```
Extract Gene IDs from GeneSetCollection and convert to characters:
```{r}
library(GSEABase)

# Extract gene names from GeneSetCollection
pi3k_genes <- geneIds(sample_tbsignatures3$HALLMARK_PI3k_akt)[[1]]
tgf_beta_genes <- geneIds(sample_tbsignatures3$HALLMARK_TGF_BETA)[[1]]
p53_genes <- geneIds(sample_tbsignatures3$HALLMARK_P53)[[1]]
WNT_BETA_CATENIN_genes <- geneIds(sample_tbsignatures3$HALLMARK_WNT_BETA_CATENIN)[[1]]

# Convert to character vectors
sample_tbsignatures3$HALLMARK_PI3k_akt <- as.character(pi3k_genes)
sample_tbsignatures3$HALLMARK_TGF_BETA <- as.character(tgf_beta_genes)
sample_tbsignatures3$HALLMARK_P53 <- as.character(p53_genes)
sample_tbsignatures3$HALLMARK_WNT_BETA_CATENIN <- as.character(WNT_BETA_CATENIN_genes)
```

Run TBSignatureProfiler:
```{r}
library(HGNChelper)

ssgsea_res <- runTBsigProfiler(
    input = se, 
    useAssay = "TPM", 
    signatures = sample_tbsignatures3,  
    algorithm = "ssGSEA",
    combineSigAndAlgorithm = TRUE,
    parallel.sz = 4,
    update_genes = FALSE
)
```

Save the SummarizedExperiment and TBSignatureProfiler run:
```{r}
saveRDS(se, file = "TB6monthSE.rds")

saveRDS(ssgsea_res, file = "TB6monthsSE_run2.rds") # change this based on what run you are doing 
```

Save row data into a table from the TBSignatureProfiler run:
```{r}
write.csv(colData(ssgsea_res), file = "gene_annotations2.csv", row.names = TRUE) # change this based on what run you are doing
```

Visualize the results with a boxplot 
```{r}
library(ggplot2)
plot_boxplot <- signatureBoxplot(inputData = ssgsea_res, 
                                 name = "Boxplots of Signatures, ssGSEA", 
                                 signatureColNames = names(sample_tbsignatures3),#TBsignatures, sample_tbsignatures
                                 annotationColName = "Timepoint", rotateLabels = FALSE, fill_colors = c("blue","orange", "green"))#,,"blue","orange", "green"))
print(plot_boxplot)
```

Save boxplot:
```{r}
ggsave("~/TBLung/TB_6Month/ssGSEA_signature_boxplot2.png", plot = plot_boxplot, width = 8, height = 6, dpi = 300)
```

Adjusted boxplot that looks better:
```{r}
library(ggplot2)
plot_boxplot2 <- plot_boxplot + labs(x = "Sample Group Timepoint", y = "Score", fill = "Timepoint") + 
  theme(
  text = element_text(size = 7),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title = element_text(size = 14),
  plot.title = element_text(size = 16, hjust = 0.5),
  plot.margin = margin(10, 10, 10, 10))
```

Save manipulated boxplot:
```{r}
ggsave(
  filename = "boxplot_MH2.png",  # Specify the filename
  plot = plot_boxplot2,              # The plot to save
  width = 10,                       # Width in inches
  height = 12,                      # Height in inches
  dpi = 300                         # Resolution in dots per inch
)
```

Make a spaghetti plot with the gene annotations: 
```{r}
# import the gene_annotations2.csv 
library(readr)
gene_annotations2 <- read_csv("gene_annotations2.csv")
class(gene_annotations2)
```

Ensure data types are correct:
```{r}
str(gene_annotations2)
```

Restructure the data to make it into tidy format so that it is better for ggplot
```{r}
library(dplyr)
library(tidyr)

# Rename the first column to "ID" and convert necessary columns to factors
gene_annotations2 <- gene_annotations2 %>%
  rename(ID = `...1`) %>%
  mutate(
    ID = as.factor(ID),
    Sampler = as.factor(Sampler),
    Timepoint = as.factor(Timepoint))

# Reshape the dataset into long format for easier visualization
tidy_data <- gene_annotations2 %>%
  pivot_longer(
    cols = starts_with("HALLMARK") | starts_with("Sweeney") | starts_with("Zak"),
    names_to = "Signature",
    values_to = "Score")
```

Making a spaghetti plot to track each of the samplers (A, B, C, D, E, F)
```{r}
library(RColorBrewer)

ggplot(tidy_data, aes(x = Timepoint, y = Score, group = Sampler, color = Sampler)) +
  geom_line(size = 1.2, alpha = 0.8) +  # Makes a line to connect the points across the three timepoints
  geom_point(size = 3) +  # plots the scores as points 
  facet_wrap(~Signature, scales = "free_y") +  # One panel for each of the six gene signatures
  scale_color_brewer(palette = "Set2") +
  theme_light() +
  labs(title = "Spaghetti Plot of Gene Signatures Across Timepoints",
       x = "Timepoint",
       y = "Score",
       color = "Sampler") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 6.5, face = "bold"))  # Rotate x-axis labels for readability and makes the plot titles bold and smaller so they all fit
```

Save the plot as .png file
```{r}
ggsave("spaghetti_plot_TB6months.png", width = 10, height = 6, dpi = 300)
```

